---
title: "Final Project"
author: "Angelika Glogov, Cynthia Kennedy, Mac Riley, & Jonathan Fromm"
format: html
editor: visual
---

## Link to animations
https://docs.google.com/document/d/1UTypArRaOsMwPB56y-8Q2uVr4tfJlxWDACOPvhng6GI/edit?usp=sharing

## Introduction

Our animated data visualizations are a tool for exploring how much household air 
pollution (HAP) differs within and between regions and how those discrepancies 
have evolved over the 2010s. HAP is often emitted from cooking with solid fuels 
such as wood or coal, smoking, cleaning products, and mold inside homes. 
Specifically, our animations seek to inform the question: How did the dispersion 
of death rates attributed to HAP (for each sex) among and within different 
regions change over the 2010s, and what does that indicate about the equality of 
regions in terms of how this global burden is shared? The animations are 
organized by six geographical divisions created by the World Health Organization 
(WHO) for research purposes: the African Region, the Region of the Americas, the 
Eastern Mediterranean Region, the European Region, the South-East Asia Region, 
and the Western Pacific Region. By illustrating changes in the spread of 
age-standardized (adjusted for differences in age distribution among countries) 
death rates attributed to HAP-induced causes between regions over this decade, 
we can more clearly identify whether regions are converging towards similar 
levels of risk. Thus, our central goal is to determine if regions are more 
equally sharing the global burden of HAP, or if disparities are increasing 
across different parts of the world.

## Data Source and Collection

For this project, we used a dataset from the World Health Organization’s Global 
Health Observatory. We sourced it from their indicator entitled “Household Air 
pollution attributable death rate (per 100,000 population, age-standardized).” 
This data estimates national mortality rates attributed to HAP-induced causes 
(per a standardized measure of 100,000 people) annually over the period 
2010-2019. These HAP-induced causes are Acute lower respiratory conditions (1), 
Trachea, bronchus, lung cancers (2), Ischaemic heart disease (3), Stroke (4), 
and Chronic Obstructive pulmonary disease (5).

This data was collected by the WHO through comparative risk assessment methods 
that estimate the number of deaths or disability-adjusted life years (DALYs) in 
each country that can be attributed to air pollution exposure (focusing on fine 
particulate matter, or PM2.5). The WHO first compiled epidemiological evidence 
that displayed an association between the exposure to air pollution and several 
diseases, including lower respiratory infections, strokes, ischaemic heart 
diseases, chronic obstructive pulmonary diseases, and lung cancers. They used 
these conclusions and an alternative baseline level of air pollution exposure 
which poses a minimum health risk (while holding other conditions constant) to 
determine the increased risk of these diseases being induced by air pollution 
exposure. Next, they combined this with information on how widespread HAP 
exposure is in a given population to calculate the population attributable 
fraction (PAF), which represents the proportion of disease burden that can be 
attributed to air pollution exposure. When applied to the total disease burden 
in the population, this fraction revealed the total number of deaths or DALYs 
linked to HAP. Lastly, the WHO converted these raw numbers into age-standardized 
rates per 100,000 population to adjust for the differences in age distribution 
and total population size across regions. The resulting data allow us to 
quantify the burden of disease caused by HAP exposure. However, the WHO 
acknowledges several limitations to the accuracy of these estimates. They 
include the current quality of epidemiological evidence, the availability of 
country-level background disease burden data, the quality of global ambient 
air pollution exposure models and ground-level pollutant measurements, and the 
timeliness of updates to WHO’s global household energy model.

Since each data point represents the age-standardized death rate attributable to 
HAP-induced causes for a given country in a specific year, the observational 
unit in this dataset is a country-year. In our analysis of the data, we used 
data points from the combined All Causes variable, which represents deaths 
attributed to each of the five HAP-induced causes per 100,000 population. We 
used these data points for the y-axes of all of our animations. We used the 
Location variable to group our data points in the bars and lines corresponding 
to different countries and regions of the world. We separated our line graph 
animations for regions using the Sex variable to explore any differences in 
HAP-induced death rates between males and females. Lastly, we used the Period 
variable to animate our visualizations (placing it on the x-axis for the line 
graph visualizations) so we could assess differences in the data over each year 
in the 2010s. 

## Relation to Current State of the Field

Since 400 BCE, when the ancient Greek physician and philosopher Hippocrates 
first detected a link between disease and air quality, air pollution has been 
recognized as a critical environmental threat to global health. PM2.5, or fine 
particulate matter, is an especially harmful pollutant because of its miniscule 
size. These tiny particles can penetrate the bloodstream and enter the lungs, 
increasing the risk of cardiovascular and respiratory diseases that can lead to 
death (“Type of pollutants,” 2025). According to the WHO, household air 
pollution disproportionately affects lower income countries, and women and 
children typically bear the greatest disease burden related to HAP exposure 
(“Household air pollution,” 2024).

This global health risk has motivated researchers to work towards quantifying 
the disease burden attributable to air pollution. Global burden-of-disease 
studies consistently identify air pollution, both ambient and household, as one 
of the leading risk factors for death around the world (McDuffie et al., 2021). 
The WHO dataset used to create this animation is focused on household air 
pollution, and its standardized estimates allow us to understand the 
distribution of the mortality burden across countries, and how it has evolved 
over time. The data is also separated by sex, which allows us to empirically 
investigate the currently held notion that women bear a greater disease burden 
related to HAP exposure than men do. By creating digestible visualizations for 
these comparisons, we can identify inequalities, evaluate global progress, and 
inform policy strategies for improving air quality and reducing health 
disparities.


```{r, echo = FALSE, warning = FALSE, message = FALSE}

suppressMessages(library(tidyverse))

library(tidyverse)
library(ggplot2)
library(gganimate)

rawdata <- read_csv("data/pollution.csv")

filtered <- rawdata |>
  filter(Dim2ValueCode == "GHECAUSE_GHE000000", Dim1ValueCode != "SEX_BTSX")
colnames(filtered)[13] <- "Gender"

continent <- filtered |>
  group_by(ParentLocation, Period, Gender) |>
  mutate(deaths = mean(FactValueNumeric))

filtered <- filtered |>
  group_by(Location, Period) |>
  mutate(deaths = mean(FactValueNumeric))

continent <- continent |>
  group_by(ParentLocation, Period) |>
  arrange(desc(ParentLocation)) |>
  group_by(ParentLocation, Period, Gender) |>
  slice(1)
  
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}

#| fig-width: 20
#| fig-height: 5

ggplot(continent, 
  aes(x = Period, y = deaths, fill = ParentLocation, color = ParentLocation)) + 
  geom_line() + 
  geom_segment(
    aes(xend = 2019, yend = deaths), 
    linetype = 2, 
    colour = "grey") + 
  geom_point(size = 2) + 
  scale_fill_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) + 
  scale_color_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) +
  scale_x_continuous(breaks = seq(2010, 2019, by = 1)) +
  geom_text(
    aes(x = 2019.1, label = ParentLocation), 
    hjust = 0) + 
  facet_wrap(~ Gender, dir = "v") + 
  gganimate::transition_reveal(Period) +
  coord_cartesian(clip = 'off') + 
  labs(
    x = "Year",
    y = "Average Deaths (per 100,000 population)",
    title = "Average HAP-Attributable Death Rates (by Continent and Sex)") + 
  theme_minimal() + 
  theme(plot.margin = margin(5.5, 100, 5.5, 5.5), 
        legend.position = "none")

```

```{r, echo = FALSE, warning = FALSE, message = FALSE}

nogender <- rawdata |>
  filter(Dim2ValueCode == "GHECAUSE_GHE000000", Dim1ValueCode == "SEX_BTSX")
colnames(nogender)[13] <- "Gender"

filtered <- rawdata |>
  filter(Dim2ValueCode == "GHECAUSE_GHE000000", Dim1ValueCode != "SEX_BTSX")
colnames(filtered)[13] <- "Gender"

filtered <- filtered |>
  group_by(Location, Period) |>
  mutate(deaths = mean(FactValueNumeric))

continent <- filtered|>
  group_by(ParentLocation, Period) |>
  arrange(desc(ParentLocation)) |>
  group_by(ParentLocation, Period, Gender)

continentnogender <- nogender |>
  group_by(ParentLocation, Period, Gender) |>
  mutate(deaths = (mean(FactValueNumeric))) |>
  group_by(ParentLocation, Period) |>
  arrange(desc(ParentLocation)) |>
  group_by(ParentLocation, Period, Gender) |>
  filter(!duplicated(deaths))

continent <- continent |>
  group_by(ParentLocation, Period) |>
  arrange(desc(ParentLocation)) |>
  group_by(ParentLocation, Period, Gender) |>
  slice(1)
  
```

```{r, echo = FALSE, warning = FALSE, message = FALSE}

ggplot(data = continentnogender, aes(x = ParentLocation, y = deaths, fill = ParentLocation)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) +
  theme(legend.position = "none") +
  transition_states(
    Period,
    transition_length = 2,
    state_length = 1) +
  enter_fade() +
  exit_shrink() +
  labs(title = 'Year: {closest_state}', x = 'Region', y = 'Average Deaths from Air Pollution (per 100,000 population)') +
  ease_aes('sine-in-out') 

```

```{r, echo = FALSE, warning = FALSE, message = FALSE}

continent <- filtered|>
  group_by(ParentLocation, Period) |>
  arrange(desc(ParentLocation)) |>
  group_by(ParentLocation, Period, Gender)

africa <- nogender |>
  filter(ParentLocation == "Africa") |>
  group_by(Location, Period, Gender) |>
  mutate(deaths = (mean(FactValueNumeric))) |>
  group_by(Location, Period) |>
  arrange(desc(Location)) |>
  group_by(Location, Period, Gender) |>
  filter(!duplicated(deaths))

mostpopulous <- africa |>
  filter(Location == "Nigeria" || Location == "Ethiopia" || Location == "Egypt" || Location == "Democratic Republic of the Congo" || Location == "United Republic of Tanzania" || Location == "South Africa" || Location == "Kenya" || Location == "Sudan" || Location == "Uganda") |>
  group_by(Location, Period, Gender) |>
  mutate(deaths = (mean(FactValueNumeric))) |>
  group_by(Location, Period) |>
  arrange(desc(Location)) |>
  group_by(Location, Period, Gender) |>
  slice(1)

```

```{r, echo = FALSE, warning = FALSE, message = FALSE}

ggplot(
  mostpopulous, 
  aes(x = Period, y = deaths, fill = Location, color = Location)) + 
  geom_line() + 
  geom_segment(
    aes(xend = 2019, yend = deaths), 
    linetype = 2, 
    colour = "grey") + 
  geom_point(size = 2) + 
  scale_x_continuous(breaks = seq(2010, 2019, by = 1)) +
  geom_text(
    aes(x = 2019.1, label = Location), 
    hjust = 0) + 
  gganimate::transition_reveal(Period) +
  coord_cartesian(clip = 'off') + 
  labs(
    x = "Year",
    y = "Average Deaths (per 100,000 population)",
    title = "Average Deaths from Air Pollution in Africa's 9 Most Populous Countries") +
  theme_minimal() + 
  theme(plot.margin = margin(5.5, 40, 5.5, 5.5), 
        legend.position = "none")

```

```{r, echo = FALSE, warning = FALSE, message = FALSE}

africa <- continent |>
  filter(ParentLocation == "Africa") |>
  group_by(Location, Period, Gender) |>
  mutate(deaths = (mean(FactValueNumeric))) |>
  group_by(Location, Period) |>
  arrange(desc(Location)) |>
  group_by(Location, Period, Gender) |>
  filter(!duplicated(deaths))

mostpopulous <- africa |>
  filter(Location == "Nigeria" || Location == "Ethiopia" || Location == "Egypt" || Location == "Democratic Republic of the Congo" || Location == "United Republic of Tanzania" || Location == "South Africa" || Location == "Kenya" || Location == "Sudan" || Location == "Uganda") |>
  group_by(Location, Period, Gender) |>
  mutate(deaths = (mean(FactValueNumeric))) |>
  group_by(Location, Period) |>
  arrange(desc(Location)) |>
  group_by(Location, Period, Gender) |>
  filter(!duplicated(deaths))

ggplot(
  mostpopulous, 
  aes(x = Period, y = deaths, fill = Location, color = Location)) + 
  geom_line() + 
  geom_segment(
    aes(xend = 2019, yend = deaths), 
    linetype = 2, 
    colour = "grey") + 
  geom_point(size = 2) + 
  #scale_fill_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) + 
  #scale_color_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) +
  scale_x_continuous(breaks = seq(2010, 2019, by = 1)) +
  geom_text(
    aes(x = 2019.1, label = Location), 
    hjust = 0) + 
  facet_wrap(~ Gender, dir = "v") + 
  gganimate::transition_reveal(Period) +
  coord_cartesian(clip = 'off') + 
  labs(
    x = "Year",
    y = "Average Deaths (per 100,000 population)",
    title = "Average Deaths from Air Pollution in Africa's 9 Most Populous Countries") + 
  theme_minimal() + 
  theme(plot.margin = margin(5.5, 40, 5.5, 5.5), 
        legend.position = "none")

```

## Results and Discussion

The results we found from our animated visualizations showed several important 
insights about how HAP has affected different regions across the globe during 
the 2010s. Our biggest takeaway was that while age-standardized death rates 
attributable to HAP declined across all six WHO regions over the decade, the 
gaps from region to region remained similar. For example, Africa was 
consistently the region with the highest mortality rate, followed by South-East 
Asia and the Western Pacific respectfully. On the other side of the spectrum, 
the Americas and Europe maintained relatively low compared to the previously 
mentioned regions and retained steadily declining rates throughout the period. 
These disparities led us to do a deeper dive into the HAP inequality between the 
nine most populous countries within Africa, showing even greater division within 
the continent itself. We saw the Democratic Republic of the Congo contributing 
significantly to the overall deaths within Africa and Uganda, Nigeria, Ethiopia, 
Kenya, and Tanzania having around 120 deaths per 100 thousand in the “middle 
ground” of the continent. South Africa was significantly lower than the majority 
of the group. These findings suggest that although global progress has been 
made, the distribution of HAP-related mortality remains very unequal.

## Implications for Research Questions and Global Health Equity

Our research question asked if regional death rates that were attributable to 
HAP were converging during the 2010s. We were looking to uncover what this says 
about equality across the world in terms of environmental burdens. The 
visualizations indicate a decline in deaths per capita but an unequal 
distribution by region. Absolute death rates have fallen across regions, which 
implies that interventions such as cleaner cooking fuels, improved ventilation, 
and global awareness campaigns may be contributing to improved health. However, 
as mentioned above, the comparatively high death rates in Africa and Asia 
demonstrate that structural problems exist in energy access, infrastructure, 
technology, and economic development. These disparities continue to shape where 
HAP remains the most lethal. Our visualization focusing on Africa reinforces 
this pattern even within a single region. While all nine of Africa’s most 
populous countries experienced declines from 2010 to 2019, the lines do not 
converge. Rather we see large gaps persist between countries like the Democratic 
Republic of the Congo, where death rates remain extremely high, and countries 
such as South Africa, which consistently reports far lower burdens. This lack of 
convergence highlights that improvements within Africa have been uneven, 
mirroring the global inequality we observe across regions. Some countries appear 
to be benefiting more rapidly from cleaner energy transitions and health 
infrastructure than others. Additionally, looking at the continent as a whole, 
the fact that Africa remained the most burdened region in every year of the 
animation suggests that global progress has not translated proportionally across 
populations, raising important questions about environmental justice and 
resource allocation. Looking into policy, these findings make clear the need for 
change with aiding the regions that take on the death burden of climate change.

## Methodological Strengths and Limitations

Our main concern with the estimates that we relied upon to create these 
visualizations were that they rely on modeled data, not direct measurement. As 
the WHO notes, these mortality estimates depend on exposure modeling and 
assumptions about the distribution of HAP within populations. Limitations in air 
quality monitoring networks, particularly in low-income countries, introduce 
uncertainty into these estimates which will likely underestimate or overestimate 
true death rates, raising the error rates for these countries. Our 
visualization, while effective for identifying broad trends, cannot capture the 
statistical uncertainty or confidence intervals around these modeled values.

## Broader Significance

Overall, our findings illustrate that global efforts to reduce household air 
pollution are yielding positive results, but not equally or uniformly. The 
steady decline in HAP-attributable mortality during the 2010s reflects real 
progress in public health, energy infrastructure, and pollution reduction 
initiatives. However, the persistent concentration of high mortality rates in 
Africa, South-East Asia, and parts of the Western Pacific highlights ongoing 
inequalities in technological access, socioeconomic conditions, and public 
health capacity. Our Africa-focused visualization further underscores this point 
by showing that even within a single region, countries are not converging toward 
a shared mortality level. Instead, stark gaps remain between nations such as the 
Democratic Republic of the Congo and South Africa, signaling that progress has 
been highly uneven.

From a global health perspective, this encapsulates the critical truth that 
reducing risk factors requires a deep investment into the regions and a 
continual emphasis on environmental justice. Our visualization makes these 
disparities visible, reminding policymakers, researchers, and the public that 
while the world is moving in the right direction, many regions remain 
disproportionately burdened by preventable environmental health risks.

## Works Cited

World Health Organization. (2024, October 16). Household air pollution and 
health. https://www.who.int/news-room/fact-sheets/detail/household-air-pollution-and-health

McDuffie, E., Martin, R., Yin, H., & Brauer, M. (2021). Global burden of disease 
from major air pollution sources: A global approach. Research Report — Health 
Effects Institute, 2021(210), 1–45. https://pmc.ncbi.nlm.nih.gov/articles/PMC9501767/

World Health Organization. (2025). Types of pollutants — Air quality & health. https://www.who.int/teams/environment-climate-change-and-health/air-quality-and-health/health-impacts/types-of-pollutants
