---
title: "Final Project"
format: html
editor: visual
---

## Introduction

Our animated data visualizations are a tool for exploring how much household air 
pollution (HAP) differs within and between regions and how those discrepancies 
have evolved over the 2010s. HAP is often emitted from cooking with solid fuels 
such as wood or coal, smoking, cleaning products, and mold inside homes. 
Specifically, our animations seek to inform the question: How did the dispersion 
of death rates attributed to HAP (for each sex) among and within different 
regions change over the 2010s, and what does that indicate about the equality of 
regions in terms of how this global burden is shared? The animations are 
organized by six geographical divisions created by the World Health Organization 
(WHO) for research purposes: the African Region, the Region of the Americas, the 
Eastern Mediterranean Region, the European Region, the South-East Asia Region, 
and the Western Pacific Region. By illustrating changes in the spread of 
age-standardized (adjusted for differences in age distribution among countries) 
death rates attributed to HAP-induced causes between regions over this decade, 
we can more clearly identify whether regions are converging towards similar 
levels of risk. Thus, our central goal is to determine if regions are more 
equally sharing the global burden of HAP, or if disparities are increasing 
across different parts of the world.

## Data Source and Collection

For this project, we used a dataset from the World Health Organization’s Global 
Health Observatory. We sourced it from their indicator entitled “Household Air 
pollution attributable death rate (per 100,000 population, age-standardized).” 
This data estimates national mortality rates attributed to HAP-induced causes 
(per a standardized measure of 100,000 people) annually over the period 
2010-2019. These HAP-induced causes are Acute lower respiratory conditions (1), 
Trachea, bronchus, lung cancers (2), Ischaemic heart disease (3), Stroke (4), 
and Chronic Obstructive pulmonary disease (5).

This data was collected by the WHO through comparative risk assessment methods 
that estimate the number of deaths or disability-adjusted life years (DALYs) in 
each country that can be attributed to air pollution exposure (focusing on fine 
particulate matter, or PM2.5). The WHO first compiled epidemiological evidence 
that displayed an association between the exposure to air pollution and several 
diseases, including lower respiratory infections, strokes, ischaemic heart 
diseases, chronic obstructive pulmonary diseases, and lung cancers. They used 
these conclusions and an alternative baseline level of air pollution exposure 
which poses a minimum health risk (while holding other conditions constant) to 
determine the increased risk of these diseases being induced by air pollution 
exposure. Next, they combined this with information on how widespread HAP 
exposure is in a given population to calculate the population attributable 
fraction (PAF), which represents the proportion of disease burden that can be 
attributed to air pollution exposure. When applied to the total disease burden 
in the population, this fraction revealed the total number of deaths or DALYs 
linked to HAP. Lastly, the WHO converted these raw numbers into age-standardized 
rates per 100,000 population to adjust for the differences in age distribution 
and total population size across regions. The resulting data allow us to 
quantify the burden of disease caused by HAP exposure. However, the WHO 
acknowledges several limitations to the accuracy of these estimates. They 
include the current quality of epidemiological evidence, the availability of 
country-level background disease burden data, the quality of global ambient 
air pollution exposure models and ground-level pollutant measurements, and the 
timeliness of updates to WHO’s global household energy model.

Since each data point represents the age-standardized death rate attributable to 
HAP-induced causes for a given country in a specific year, the observational 
unit in this dataset is a country-year. In our analysis of the data, we used 
data points from the combined All Causes variable, which represents deaths 
attributed to each of the five HAP-induced causes per 100,000 population. We 
used these data points for the y-axes of all of our animations. We used the 
Location variable to group our data points in the bars and lines corresponding 
to different countries and regions of the world. We separated our line graph 
animations for regions using the Sex variable to explore any differences in 
HAP-induced death rates between males and females. Lastly, we used the Period 
variable to animate our visualizations (placing it on the x-axis for the line 
graph visualizations) so we could assess differences in the data over each year 
in the 2010s. 

## Relation to Current State of the Field

Since 400 BCE, when the ancient Greek physician and philosopher Hippocrates 
first detected a link between disease and air quality, air pollution has been 
recognized as a critical environmental threat to global health. PM2.5, or fine 
particulate matter, is an especially harmful pollutant because of its miniscule 
size. These tiny particles can penetrate the bloodstream and enter the lungs, 
increasing the risk of cardiovascular and respiratory diseases that can lead to 
death (“Type of pollutants,” 2025). According to the WHO, household air 
pollution disproportionately affects lower income countries, and women and 
children typically bear the greatest disease burden related to HAP exposure 
(“Household air pollution,” 2024).

This global health risk has motivated researchers to work towards quantifying 
the disease burden attributable to air pollution. Global burden-of-disease 
studies consistently identify air pollution, both ambient and household, as one 
of the leading risk factors for death around the world (McDuffie et al., 2021). 
The WHO dataset used to create this animation is focused on household air 
pollution, and its standardized estimates allow us to understand the 
distribution of the mortality burden across countries, and how it has evolved 
over time. The data is also separated by sex, which allows us to empirically 
investigate the currently held notion that women bear a greater disease burden 
related to HAP exposure than men do. By creating digestible visualizations for 
these comparisons, we can identify inequalities, evaluate global progress, and 
inform policy strategies for improving air quality and reducing health 
disparities.


```{r, echo = FALSE, warning = FALSE, message=FALSE}

suppressMessages(library(tidyverse))

library(tidyverse)
library(ggplot2)
library(gganimate)

rawdata <- read_csv("data/pollution.csv")

filtered <- rawdata |>
  filter(Dim2ValueCode == "GHECAUSE_GHE000000", Dim1ValueCode != "SEX_BTSX")
colnames(filtered)[13] <- "Gender"

continent <- filtered |>
  group_by(ParentLocation, Period, Gender) |>
  mutate(deaths = mean(FactValueNumeric))

filtered <- filtered |>
  group_by(Location, Period) |>
  mutate(deaths = mean(FactValueNumeric))

continent <- continent |>
  group_by(ParentLocation, Period) |>
  arrange(desc(ParentLocation)) |>
  group_by(ParentLocation, Period, Gender) |>
  slice(1)
  
```

```{r, echo = FALSE, warning = FALSE, message=FALSE}

#| fig-width: 20
#| fig-height: 5

ggplot(
  continent, 
  aes(x = Period, y = deaths, fill = ParentLocation, color = ParentLocation)) + 
  geom_line() + 
  geom_segment(
    aes(xend = 2019, yend = deaths), 
    linetype = 2, 
    colour = "grey") + 
  geom_point(size = 2) + 
  scale_fill_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) + 
  scale_color_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) +
  scale_x_continuous(breaks = seq(2010, 2019, by = 1)) +
  geom_text(
    aes(x = 2019.1, label = ParentLocation), 
    hjust = 0) + 
  facet_wrap(~ Gender, dir = "v") + 
  gganimate::transition_reveal(Period) +
  coord_cartesian(clip = 'off') + 
  labs(
    x = "Year",
    y = "Average Deaths (per 100,000 population)",
    title = "Average HAP-Attributable Death Rates (by Continent and Sex)") + 
  theme_minimal() + 
  theme(plot.margin = margin(5.5, 100, 5.5, 5.5), 
        legend.position = "none")

```

```{r, echo = FALSE, warning = FALSE, message=FALSE}

nogender <- rawdata |>
  filter(Dim2ValueCode == "GHECAUSE_GHE000000", Dim1ValueCode == "SEX_BTSX")
colnames(nogender)[13] <- "Gender"

filtered <- rawdata |>
  filter(Dim2ValueCode == "GHECAUSE_GHE000000", Dim1ValueCode != "SEX_BTSX")
colnames(filtered)[13] <- "Gender"

filtered <- filtered |>
  group_by(Location, Period) |>
  mutate(deaths = mean(FactValueNumeric))

continent <- filtered|>
  group_by(ParentLocation, Period) |>
  arrange(desc(ParentLocation)) |>
  group_by(ParentLocation, Period, Gender)

continentnogender <- nogender |>
  group_by(ParentLocation, Period, Gender) |>
  mutate(deaths = (mean(FactValueNumeric))) |>
  group_by(ParentLocation, Period) |>
  arrange(desc(ParentLocation)) |>
  group_by(ParentLocation, Period, Gender) |>
  filter(!duplicated(deaths))

continent <- continent |>
  group_by(ParentLocation, Period) |>
  arrange(desc(ParentLocation)) |>
  group_by(ParentLocation, Period, Gender) |>
  slice(1)
  
```

```{r, echo = FALSE, warning = FALSE, message=FALSE}

ggplot(data = continentnogender, aes(x = ParentLocation, y = deaths, fill = ParentLocation)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) +
  theme(legend.position = "none") +
  # Here comes the gganimate code
  transition_states(
    Period,
    transition_length = 2,
    state_length = 1
  ) +
  enter_fade() +
  exit_shrink() +
  labs(title = 'Year: {closest_state}', x = 'Region', y = 'Average Deaths from Air Pollution (per 100,000 population)') +
  ease_aes('sine-in-out') 

```

```{r, echo = FALSE, warning = FALSE, message=FALSE}

continent <- filtered|>
  group_by(ParentLocation, Period) |>
  arrange(desc(ParentLocation)) |>
  group_by(ParentLocation, Period, Gender)

africa <- nogender |>
  filter(ParentLocation == "Africa") |>
  group_by(Location, Period, Gender) |>
  mutate(deaths = (mean(FactValueNumeric))) |>
  group_by(Location, Period) |>
  arrange(desc(Location)) |>
  group_by(Location, Period, Gender) |>
  filter(!duplicated(deaths))

mostpopulous <- africa |>
  filter(Location == "Nigeria" || Location == "Ethiopia" || Location == "Egypt" || Location == "Democratic Republic of the Congo" || Location == "United Republic of Tanzania" || Location == "South Africa" || Location == "Kenya" || Location == "Sudan" || Location == "Uganda") |>
  group_by(Location, Period, Gender) |>
  mutate(deaths = (mean(FactValueNumeric))) |>
  group_by(Location, Period) |>
  arrange(desc(Location)) |>
  group_by(Location, Period, Gender) |>
  slice(1)

```

```{r, echo = FALSE, warning = FALSE, message=FALSE}

ggplot(
  mostpopulous, 
  aes(x = Period, y = deaths, fill = Location, color = Location)) + 
  geom_line() + 
  geom_segment(
    aes(xend = 2019, yend = deaths), 
    linetype = 2, 
    colour = "grey") + 
  geom_point(size = 2) + 
  #scale_fill_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) + 
  #scale_color_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) +
  scale_x_continuous(breaks = seq(2010, 2019, by = 1)) +
  geom_text(
    aes(x = 2019.1, label = Location), 
    hjust = 0) + 
  gganimate::transition_reveal(Period) +
  coord_cartesian(clip = 'off') + 
  labs(
    x = "Year",
    y = "Average Deaths (per 100,000 population)",
    title = "Average Deaths from Air Pollution in Africa's 9 Most Populous Countries") +
  theme_minimal() + 
  theme(plot.margin = margin(5.5, 40, 5.5, 5.5), 
        legend.position = "none")

```

```{r, echo = FALSE, warning = FALSE, message=FALSE}

africa <- continent |>
  filter(ParentLocation == "Africa") |>
  group_by(Location, Period, Gender) |>
  mutate(deaths = (mean(FactValueNumeric))) |>
  group_by(Location, Period) |>
  arrange(desc(Location)) |>
  group_by(Location, Period, Gender) |>
  filter(!duplicated(deaths))

mostpopulous <- africa |>
  filter(Location == "Nigeria" || Location == "Ethiopia" || Location == "Egypt" || Location == "Democratic Republic of the Congo" || Location == "United Republic of Tanzania" || Location == "South Africa" || Location == "Kenya" || Location == "Sudan" || Location == "Uganda") |>
  group_by(Location, Period, Gender) |>
  mutate(deaths = (mean(FactValueNumeric))) |>
  group_by(Location, Period) |>
  arrange(desc(Location)) |>
  group_by(Location, Period, Gender) |>
  filter(!duplicated(deaths))

ggplot(
  mostpopulous, 
  aes(x = Period, y = deaths, fill = Location, color = Location)) + 
  geom_line() + 
  geom_segment(
    aes(xend = 2019, yend = deaths), 
    linetype = 2, 
    colour = "grey") + 
  geom_point(size = 2) + 
  #scale_fill_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) + 
  #scale_color_manual(values = c("#3A90D6", "#3A51D6", "#5E3AD6", "#853AD6", "#AF3AD6", "#D63ACC")) +
  scale_x_continuous(breaks = seq(2010, 2019, by = 1)) +
  geom_text(
    aes(x = 2019.1, label = Location), 
    hjust = 0) + 
  facet_wrap(~ Gender, dir = "v") + 
  gganimate::transition_reveal(Period) +
  coord_cartesian(clip = 'off') + 
  labs(
    x = "Year",
    y = "Average Deaths (per 100,000 population)",
    title = "Average Deaths from Air Pollution in Africa's 9 Most Populous Countries") + 
  theme_minimal() + 
  theme(plot.margin = margin(5.5, 40, 5.5, 5.5), 
        legend.position = "none")

```
